---
title: "Cross-validated correlation analysis"
date: "`r format(Sys.time(), '%d %B, %Y')`"
classoption: "aspectratio=169"
author: "Yongjin P Park"
output:
    beamer_presentation:
        colortheme: "orchid"
        keep_tex: true
        latex_engine: xelatex
        slide_level: 2
---

```{r include = FALSE}
library(data.table)
library(tidyverse)
library(patchwork)
theme_set(theme_classic() +
          theme(axis.text = element_text(size=5)) +
          theme(legend.text = element_text(size=4)) +
          theme(legend.key.size = unit(.3, "lines")) +
          theme(legend.title = element_text(size=6)))
source("Util.R")
source("Setup.R", )
setup.env("Fig/2a_cross_validation/")
```

## Cross validate expression polygenic scores

```{r include = FALSE}
data.dir <- "~/Dropbox/Writing/AD430/1.Results/3.eQTL/"
.files <- list.files(paste0(data.dir, "/heritability"), full.names = T)
data.dt <- do.call(rbind, lapply(.files, fread))
data.dt <- na.omit(data.dt)
```

Goal: Predict individual-level scores given summary statistics z-score vector. $G$ is $n \times p$ genotype dosage matrix.

Given the singular value decomposition results of a genotype matrix $G$, namely,

$$n^{-1/2} G = U D V^{\top},$$

we can estimate polygenic scores for the $n$ individuals:

$$\hat{\mathbf{y}} = U (D + \sqrt{\lambda}I)^{-1} V^{\top}\mathbf{z}$$

using $p$-vector $\mathbf{z}$. We also optimized the number of SVD components.

## Finding optimal number of principal components with CV cutoff 0.1

```{r include = FALSE}
rr.cutoff <- 0.1

data.dt[, pv := pnorm(`rr`/`rr.se`, lower.tail=F)]
data.dt[, fwer := p.adjust(`pv`), by = .(`celltype`, `nPC`)]

count.dt <-
    data.dt[`rr` > rr.cutoff & `fwer` < .05,
            .(nmodels = .N,
              ngenes = length(unique(`gene`))),
            by = .(nPC)] %>%
    arrange(`nPC`)
```

```{r}
plot.count <- function(count.dt){
    p1 <-
        ggplot(count.dt, aes(nPC, ngenes)) +
        geom_smooth(se=F, color="gray70") +
        geom_vline(xintercept = 37, lty=2, col=2) +
        geom_point() +
        xlab("# Principal Components") +
        ylab(paste0("# genes with CV cor > ", rr.cutoff))
    p2 <-
        ggplot(count.dt, aes(nPC, nmodels)) +
        geom_smooth(se=F, color="gray70") +
        geom_vline(xintercept = 37, lty=2, col=2) +
        geom_point() +
        xlab("# Principal Components") +
        ylab(paste0("# gene+cts with CV cor > ", rr.cutoff))
    p1|p2
}
```

```{r Fig_cv_corr_ngenes, fig.width=5, fig.height=3}
plot.count(count.dt)
```

## Histograms across different cell types

```{r Fig_cv_density, fig.width=5.5, fig.height=2.8}
.dt <- fread(paste0(data.dir, "heritability/ld_heritability_PC37.txt.gz"))
.dt[, pv := pnorm(`rr`/`rr.se`, lower.tail=F)]
.dt[, fwer := p.adjust(`pv`), by = .(`celltype`, `nPC`)]

ggplot(.dt, aes(rr)) +
    xlab("Cross validation correlation") +
    theme(strip.text = element_text(size=5)) +
    theme(axis.text = element_text(size=6)) +
    facet_wrap( ~ celltype, nrow = 3) +
    geom_histogram() +
    geom_vline(xintercept = 0.1, linewidth=.5, lty=2, col=2)
```

## Cell type specificity?

```{r Fig_cv_celltype_count, fig.width=5.5, fig.height=2.8}
rr.cutoff <- 0.1
.sum <- .dt[rr > rr.cutoff & `fwer` < .05, .(.N), by = .(celltype)]
.avg <- mean(.sum$N)

ggplot(.sum, aes(celltype, `N`)) +
    theme(axis.text.x = element_text(angle=90, vjust=1, hjust=1)) +
    geom_bar(stat="identity") +
    geom_hline(yintercept = round(.avg), lty=2, col=2)
```

## Heritability vs CV correlation


```{r Fig_cv_pve, fig.width=5, fig.height=3}
ggplot(.dt, aes(`rr`, log10(`pve`))) +
    stat_density2d_filled() +
    scale_y_continuous("PVE (heritability %)", labels = function(x) round(100*10^x)) +
    xlab("cross-validated correlation")
```

## Cell type-specific gene set enrichment analysis

```{r run_gene_set_enrichment, include = FALSE}
.dt <- fread(paste0(data.dir, "heritability/ld_heritability_PC37.txt.gz"))

.feat <-
    .dt[order(-rr), head(.SD, 1), by = .(gene, celltype)] %>%
    dcast(gene ~ celltype,
          value.var = "rr",
          fill = NA,
          fun.aggregate = max)

Z <- as.matrix(.feat[,-1])
colnames(Z) <- colnames(.feat)[-1]
rownames(Z) <- unlist(.feat[,1])

.file <- ".2a_cross_validation_gsea.txt.gz"

if.needed(.file, {

    gsea.dt <- data.table()

    make.gs.lol <- function(.dt) {
        .dt <- as.data.table(.dt) %>% unique()
        .list <-
            .dt[, .(gene = .(gene_symbol)), by = .(gs_name)] %>%
            as.list()
        .names <- .list$gs_name
        .ret <- .list$gene
        names(.ret) <- .names
        return(.ret)
    }

    gs.db <-
        msigdbr::msigdbr(species = "human", category = "C5") %>%
        dplyr::select(gs_name, gene_symbol)

    gs.lol <- make.gs.lol(gs.db)

    for(j in 1:ncol(Z)){

        .ct <- colnames(Z)[j]
        .scores <- unlist(scale(Z[, j]))
        names(.scores) <- rownames(Z)
        .scores <- .scores[!is.na(.scores)]

        .gsea <- fgsea::fgsea(pathways = gs.lol,
                              stats = .scores,
                              minSize = 50,
                              maxSize = 1000,
                              scoreType = "pos")

        .gsea <- .gsea[, .(pathway, pval, padj, size)]
        .gsea[, celltype := .ct]
        .gsea[, cat := "GO"]
        gsea.dt <- rbind(gsea.dt, .gsea)
    }

    .sub <- c("CP:KEGG","CP:BIOCARTA","CP:PID","CP:REACTOME","CP:WIKIPATHAYS")

    gs.db <-
        msigdbr::msigdbr(species = "human", category = "C2") %>%
        dplyr::filter(gs_subcat %in% .sub) %>% 
        dplyr::select(gs_name, gene_symbol)

    gs.lol <- make.gs.lol(gs.db)

    for(j in 1:ncol(Z)){

        .ct <- colnames(Z)[j]
        .scores <- scale(Z[, j])
        names(.scores) <- rownames(Z)

        .gsea <- fgsea::fgsea(pathways = gs.lol,
                              stats = .scores,
                              minSize = 50,
                              maxSize = 1000,
                              scoreType = "pos")

        .gsea <- .gsea[, .(pathway, pval, padj, size)]
        .gsea[, celltype := .ct]
        .gsea[, cat := "CAN"]
        gsea.dt <- rbind(gsea.dt, .gsea)
    }

    fwrite(gsea.dt, .file)
})
gsea.dt <- fread(.file)
```

```{r Fig_gsea_glial, fig.width=5, fig.height=3}
.dt <- gsea.dt[!str_starts(`celltype`, "Exc") & !str_starts(`celltype`, "Inh")]

.path.count <- .dt[padj < .1 & cat == "CAN", .(.N), by = .(pathway)]

max.celltype <- 3
max.per.celltype <- 4

.show <- unique(.dt[`pathway` %in% .path.count[`N` < max.celltype]$pathway,][order(pval, -size), head(.SD, max.per.celltype), by=.(celltype)][, .(pathway)])

.dt <- .dt[pathway %in% unlist(.show), ] %>%
    mutate(col = str_sub(str_remove(`pathway`,"REACTOME_"),1,50),
           row = `celltype`,
           weight = -log10(padj)) %>%
    order.pair(ret.tab = T)

ggplot(.dt, aes(row, col, fill=pmin(weight, 2))) +
    geom_tile(linewidth=.1,color="black") + xlab("cell types") + ylab("pathways") +
    theme(axis.text.x = element_text(angle=30, vjust=1, hjust=1)) +
    scale_fill_distiller("adj p-value",
                         labels = function(x) num.sci(10^(-x)),
                         palette = "RdPu", direction = 1)
```
	
## GO enrichment for glial cells

```{r Fig_go_glial, fig.width=5, fig.height=3}
.dt <- gsea.dt[!str_starts(`celltype`, "Exc") & !str_starts(`celltype`, "Inh")]

.path.count <- .dt[padj < .1 & cat == "GO" & !str_starts(`pathway`, "HP_"), .(.N), by = .(pathway)]

max.celltype <- 3
max.per.celltype <- 3

.show <- unique(.dt[`pathway` %in% .path.count[`N` < max.celltype]$pathway,][order(pval, -size), head(.SD, max.per.celltype), by=.(celltype)][, .(pathway)])

.dt <- .dt[pathway %in% unlist(.show), ] %>%
    mutate(col = str_sub(`pathway`,1,50), row = `celltype`, weight = -log10(padj)) %>%
    order.pair(ret.tab = T)

ggplot(.dt, aes(row, col, fill=pmin(weight, 2))) +
    geom_tile(linewidth=.1,color="black") + xlab("cell types") + ylab("pathways") +
    theme(axis.text.x = element_text(angle=90, vjust=1, hjust=1)) +
    scale_fill_distiller("adj p-value",
                         labels = function(x) num.sci(10^(-x)),
                         palette = "RdPu", direction = 1)
```

## GO enrichment for excitatory neurons

```{r Fig_go_exc, fig.width=5.5, fig.height=3}
.dt <- gsea.dt[str_starts(`celltype`, "Exc")]

.path.count <- .dt[padj < .05 & cat == "GO" & !str_starts(`pathway`, "HP_"), .(.N), by = .(pathway)]

max.celltype <- 3
max.per.celltype <- 3

.show <- unique(.dt[`pathway` %in% .path.count[`N` < max.celltype]$pathway,][order(pval, -size), head(.SD, max.per.celltype), by=.(celltype)][, .(pathway)])

.dt <- .dt[pathway %in% unlist(.show), ] %>%
    mutate(col = str_sub(`pathway`,1,50), row = `celltype`, weight = -log10(padj)) %>%
    order.pair(ret.tab = T)

ggplot(.dt, aes(row, col, fill=pmin(weight, 2))) +
    geom_tile(linewidth=.1,color="black") + xlab("cell types") + ylab("pathways") +
    theme(axis.text.x = element_text(angle=30, vjust=1, hjust=1)) +
    scale_fill_distiller("adj p-value",
                         labels = function(x) num.sci(10^(-x)),
                         palette = "RdPu", direction = 1)
```

## GO enrichment for inhibitory neurons

```{r Fig_go_inh, fig.width=4.5, fig.height=2.5}
.dt <- gsea.dt[str_starts(`celltype`, "Inh")]

.path.count <- .dt[padj < .1 & cat == "GO" & !str_starts(`pathway`, "HP_"), .(.N), by = .(pathway)]

max.celltype <- 2
max.per.celltype <- 3

.show <- unique(.dt[`pathway` %in% .path.count[`N` < max.celltype]$pathway,][order(pval, -size), head(.SD, max.per.celltype), by=.(celltype)][, .(pathway)])

.dt <- .dt[pathway %in% unlist(.show), ] %>%
    mutate(col = str_sub(`pathway`,1,50), row = `celltype`, weight = -log10(padj)) %>%
    order.pair(ret.tab = T)

ggplot(.dt, aes(row, col, fill=pmin(weight, 2))) +
    geom_tile(linewidth=.1,color="black") + xlab("cell types") + ylab("pathways") +
    theme(axis.text.x = element_text(angle=35, vjust=1, hjust=1)) +
    scale_fill_distiller("adj p-value",
                         labels = function(x) num.sci(10^(-x)),
                         palette = "RdPu", direction = 1)
```

## Finding gene modules by CV patterns

```{r prepare_gene_features_for_clustering}
.dt <- fread(paste0(data.dir, "heritability/ld_heritability_PC37.txt.gz"))

.feat <-
    .dt[rr > 0, .(gene, celltype, rr)][order(-rr), head(.SD,1), by = .(gene, celltype)] %>%
    dcast(gene ~ celltype, value.var = "rr", fill = 0, fun.aggregate = mean)

## Z <- t(apply(as.matrix(.feat[, -1]), 1, scale))
Z <- as.matrix(.feat[, -1])
colnames(Z) <- colnames(.feat[, -1])
Z[is.na(Z)] <- 0
rownames(Z) <- unlist(.feat[,1])
```

```{r cluster_by_knn}
knn <- 50
knn.file <- ".2a_cross_validation_knn_" %&% knn %&% ".rds"

if.needed(knn.file, {
    A <- build.annoy.graph(Z, knn)
    saveRDS(A, file=knn.file)
})

clust.file <- ".2a_cross_validation_cluster_" %&% knn %&% ".txt.gz"

if.needed(clust.file, {
    clust <- louvain.umap(readRDS(file=knn.file))
    fwrite(clust, clust.file)
})
clust <- fread(clust.file)
clust.sz <- clust[, .(.N), by = .(membership)]
```

```{r Fig_gene_clust_umap, fig.width=5.5, fig.height=3}
.dt <- clust[`membership` %in% clust.sz[`N` > 100]$membership]
.dt.med <- .dt[, .(V1 = median(V1), V2 = median(V2)), by = .(membership)]
.K <- length(unique(.dt$membership))

ggplot(.dt, aes(V1, V2, colour=as.factor(membership))) +
    ggrastr::rasterise(geom_point(stroke = 0, size = .8), dpi=300) +
    ggrepel::geom_text_repel(aes(label=membership), data=.dt.med, size=4, color="black") +
    scale_colour_manual("gene\ncluster",values = sample.colour(.K)) +
    theme_void() +
    theme(legend.text = element_text(size=5)) +
    theme(legend.key.size = unit(.1, "lines")) +
    xlab("UMAP1") + ylab("UMAP2")
```


## Q/C by CV confirms the cell type specificity of cis-regulatory mechanisms

```{r}
feat.dt <- clust[, .(gene, membership)] %>% left_join(.feat)
feat.melt <- melt(feat.dt, id.vars = c("gene", "membership"),
                  variable.name = "celltype",
                  value.name = "rr")
feat.celltype <-
    feat.melt[, .(rr.mu = mean(rr), rr.sd = sd(rr), nn = .N),
              by = .(membership, celltype)]

.df <- feat.celltype %>%
    mutate(row = celltype, col = membership, weight = rr.mu) %>% 
    order.pair(ret.tab = T)
```

```{r Fig_gene_clust_ct, fig.width=4.5, fig.height=3}
ggplot(.df[`nn` >= 100], aes(col, row, fill = pmin(weight, .2))) +
    xlab("cluster") + ylab("cell type") +
    geom_tile(linewidth=.1, color="gray10") +
    scale_fill_distiller("average\nCV\ncorrelation", direction=1)
```

## Gene cluster sizes

```{r Fig_gene_clust_count, fig.width=3.5, fig.height=2}
feat.celltype[, col := factor(`membership`, sort(unique(.df$col)))]
.dt.temp <- unique(feat.celltype[`nn` > 100, .(col, nn)])
ggplot(.dt.temp, aes(x = col, y = nn)) +
    geom_bar(stat="identity", width=.5) +
    ggrepel::geom_text_repel(aes(label=num.int(`nn`)), size=2) +
    xlab("cluster") + ylab("#genes")
```

