---
title: "Cross-validated correlation analysis"
date: "`r format(Sys.time(), '%d %B, %Y')`"
classoption: "aspectratio=169"
author: "Yongjin P Park"
output:
    beamer_presentation:
        colortheme: "orchid"
        keep_tex: true
        latex_engine: xelatex
        slide_level: 2
---

```{r include = FALSE}
library(data.table)
library(tidyverse)
library(patchwork)
theme_set(theme_classic())
source("Util.R")
source("Setup.R", )
setup.env("Fig/2a_cross_validation/")
```

## Cross validate expression polygenic scores

```{r include = FALSE}
data.dir <- "~/Dropbox/Writing/AD430/1.Results/3.eQTL/"
.files <- list.files(paste0(data.dir, "/heritability"), full.names = T)
data.dt <- do.call(rbind, lapply(.files, fread))
data.dt <- na.omit(data.dt)
```

Goal: Predict individual-level scores given summary statistics z-score vector. $G$ is $n \times p$ genotype dosage matrix.

Given the singular value decomposition results of a genotype matrix $G$, namely,

$$n^{-1/2} G = U D V^{\top},$$

we can estimate polygenic scores for the $n$ individuals:

$$\hat{\mathbf{y}} = U (D^{-1} + \lambda) V^{\top}\mathbf{z}$$

using $p$-vector $\mathbf{z}$. We also optimized the number of SVD components.

## Finding optimal number of principal components with CV cutoff 0.1


```{r include = FALSE}
rr.cutoff <- 0.1

count.dt <-
    data.dt[`rr` > rr.cutoff & `rr` > `rr.se` * 2,
            .(nmodels = .N,
              ngenes = length(unique(`gene`))),
            by = .(nPC)] %>%
    arrange(`nPC`)
```

```{r}
plot.count <- function(count.dt){
    p1 <-
        ggplot(count.dt, aes(nPC, ngenes)) +
        geom_smooth(se=F, color="gray70") +
        geom_vline(xintercept = 37, lty=2, col=2) +
        geom_point() +
        xlab("# Principal Components") +
        ylab(paste0("# genes with CV cor > ", rr.cutoff))
    p2 <-
        ggplot(count.dt, aes(nPC, nmodels)) +
        geom_smooth(se=F, color="gray70") +
        geom_vline(xintercept = 37, lty=2, col=2) +
        geom_point() +
        xlab("# Principal Components") +
        ylab(paste0("# gene+cts with CV cor > ", rr.cutoff))
    p1|p2
}
```

```{r Fig_cv_corr_ngenes, fig.width=5, fig.height=3}
plot.count(count.dt)
```

## Finding optimal number of principal components with CV cutoff 0.2

```{r include = FALSE}
rr.cutoff <- 0.2

count.dt <-
    data.dt[`rr` > rr.cutoff & `rr` > `rr.se` * 2,
            .(nmodels = .N,
              ngenes = length(unique(`gene`))),
            by = .(nPC)] %>%
    arrange(`nPC`)
```

```{r Fig_cv_corr_ngenes_strict, fig.width=5, fig.height=3}
plot.count(count.dt)
```

## Histograms across different cell types

```{r Fig_cv_density, fig.width=5.5, fig.height=2.8}
.dt <- fread(paste0(data.dir, "heritability/ld_heritability_PC37.txt.gz"))

ggplot(.dt, aes(rr)) +
    xlab("Cross validation correlation") +
    theme(strip.text = element_text(size=5)) +
    theme(axis.text = element_text(size=6)) +
    facet_wrap( ~ celltype, nrow = 3) +
    geom_histogram() +
    geom_vline(xintercept = 0.1, linewidth=.5, lty=2, col=2)
```

## Cell type specificity?

```{r Fig_cv_celltype_count, fig.width=5.5, fig.height=2.8}
rr.cutoff <- 0.1
.sum <- .dt[rr > rr.cutoff, .(.N), by = .(celltype)]
.avg <- mean(.sum$N)

ggplot(.sum, aes(celltype, `N`)) +
    theme(axis.text.x = element_text(angle=90, vjust=1, hjust=1)) +
    geom_bar(stat="identity") +
    geom_hline(yintercept = round(.avg), lty=2, col=2)
```

## Heritability vs CV correlation


```{r Fig_cv_pve, fig.width=5, fig.height=3}
ggplot(.dt, aes(`rr`, log10(`pve`))) +
    stat_density2d_filled() +
    scale_y_continuous("PVE (heritability %)", labels = function(x) round(100*10^x)) +
    xlab("cross-validated correlation")
```

## Finding gene modules by CV patterns

```{r prepare_gene_features_for_clustering}
.feat <-
    .dt[rr > 0, .(gene, celltype, rr)][order(-rr), head(.SD,1), by = .(gene, celltype)] %>%
    dcast(gene ~ celltype, value.var = "rr", fill = 0, fun.aggregate = mean)

Z <- t(apply(as.matrix(.feat[, -1]), 1, scale))
Z[is.na(Z)] <- 0
rownames(Z) <- unlist(.feat[,1])
```

```{r cluster_by_knn}
knn <- 50
knn.file <- ".2a_cross_validation_knn_" %&% knn %&% ".rds"

if.needed(knn.file, {
    A <- build.annoy.graph(Z, knn)
    saveRDS(A, file=knn.file)
})

clust.file <- ".2a_cross_validation_cluster_" %&% knn %&% ".txt.gz"

if.needed(clust.file, {
    clust <- louvain.umap(readRDS(file=knn.file))
    fwrite(clust, clust.file)
})
```


